---
title: "M0_FOCEI_estimation_setting_QC"
author: "Yizhen Lai"
date: "2026-06-18"
output: html_document
---

############################################################
# Simple QC before formal PopPK modeling
# Compare:
# 1) FOCEi + nlminb + 1 thread + optExpression TRUE
# 2) FOCEi + bobyqa + 1 thread + optExpression FALSE
############################################################

```{r}

rm(list = ls())

while (!is.null(dev.list())) {
  dev.off()
}

gc()

library(readr)
library(dplyr)
library(tibble)
library(nlmixr2)
library(rxode2)
library(here)

output_dir <- here("Warfarin_popPK_modeling","M0_FOCEI_estimation_setting_QC")

dir.create(
  output_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

try(rxode2::rxClean(), silent = TRUE)
try(rxode2::rxForget(), silent = TRUE)

gc()

rxode2::setRxThreads(1)

cat("rxode2 threads:", rxode2::getRxThreads(), "\n")

```

```{r}

############################################################
# 1. Read and prepare data
############################################################

dat_raw <- read_csv(
  here("Data", "warfarin_full_dataset_ID9_cleaned_for_popPKPD.csv"),
  show_col_types = FALSE
)


pk_data <- dat_raw %>%
  filter(evid == 1 | (evid == 0 & dvid == "cp")) %>%
  mutate(
    ID   = as.integer(id),
    TIME = as.numeric(time),
    DV   = ifelse(evid == 1, NA_real_, as.numeric(dv)),
    AMT  = ifelse(evid == 1, as.numeric(amt), 0),
    EVID = as.integer(evid),
    CMT  = ifelse(EVID == 1, "depot", "central")
  ) %>%
  select(ID, TIME, DV, AMT, EVID, CMT) %>%
  arrange(ID, TIME, desc(EVID))

names(pk_data)
table(pk_data$EVID, pk_data$CMT)

############################################################
# 2. Define base one-compartment proportional error model
############################################################

baseM1_prop <- function() {
  ini({
    tka <- log(1)
    tcl <- log(0.145)
    tv  <- log(8.882)

    eta.ka ~ 0.1
    eta.cl ~ 0.1
    eta.v  ~ 0.1

    prop.err <- 0.2
  })

  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)

    d/dt(depot)   = -ka * depot
    d/dt(central) =  ka * depot - cl / v * central

    cp = central / v

    cp ~ prop(prop.err)
  })
}

############################################################
# 3. Fit 1: FOCEi + nlminb
############################################################

fit_FOCEI_NLMINB_1T <- nlmixr2(
  baseM1_prop,
  pk_data,
  est = "focei",
  control = foceiControl(
    outerOpt = "nlminb",
    optExpression = TRUE
  )
)

############################################################
# 4. Fit 2: FOCEi + bobyqa
############################################################

fit_FOCEI_BOBYQA_1T <- nlmixr2(
  baseM1_prop,
  pk_data,
  est = "focei",
  control = foceiControl(
    outerOpt = "bobyqa",
    optExpression = FALSE
  )
)

```

```{r}

############################################################
# Simple QC result comparison only
# Compare FOCEi-nlminb vs FOCEi-bobyqa
############################################################

library(dplyr)
library(tibble)
library(readr)

qc_compare <- tibble(
  model_code = c(
    "FOCEI_NLMINB_1T",
    "FOCEI_BOBYQA_1T"
  ),
  method = c(
    "FOCEi",
    "FOCEi"
  ),
  setting = c(
    "1 thread + nlminb + optExpression TRUE",
    "1 thread + bobyqa + optExpression FALSE"
  ),
  message = c(
    paste(fit_FOCEI_NLMINB_1T$message, collapse = "; "),
    paste(fit_FOCEI_BOBYQA_1T$message, collapse = "; ")
  ),
  OBJF = c(
    fit_FOCEI_NLMINB_1T$objf,
    fit_FOCEI_BOBYQA_1T$objf
  ),
  AIC = c(
    AIC(fit_FOCEI_NLMINB_1T),
    AIC(fit_FOCEI_BOBYQA_1T)
  ),
  BIC = c(
    BIC(fit_FOCEI_NLMINB_1T),
    BIC(fit_FOCEI_BOBYQA_1T)
  ),
  role = c(
    "FOCEi setting comparison",
    "FOCEi setting comparison; recommended formal setting"
  )
)

print(qc_compare, n = Inf, width = Inf)

write_csv(
  qc_compare,
  file.path(output_dir, "M0_FOCEI_estimation_setting_QC_results.csv")
)

```

## QC conclusion

Before formal model development, a QC comparison of FOCEi estimation settings was conducted using the base one-compartment model with proportional residual error. Under single-thread execution, the FOCEi-nlminb setting completed successfully but returned false convergence (8), indicating suboptimal optimization stability. In contrast, the FOCEi-bobyqa setting completed with Normal exit from bobyqa and produced lower OBJF, AIC, and BIC values than the nlminb setting. Therefore, the single-thread FOCEi-bobyqa setting with optExpression = FALSE was selected as the fixed estimation setting for subsequent formal PopPK model development.


## To check the detailed result of the two model, run

print(fit_FOCEI_NLMINB_1T)
print(fit_FOCEI_BOBYQA_1T)
